### 2014-03-25 reanalysis/Fadi_Towfic_generic_medicine/towfic_et_al.rmd
mer

This document has five main parts
- Getting and formatting the data
- Reproduce some of the results to show that we are working on the same data and analysis workflow
- Remove the use of ComBat and perform an similar analysis with alternative established tools
- Estimate the error introduced by ComBat and the consequences for the conclusion of the study
- Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is mainly introduced by ComBat

```{r runparameters, results='hide', echo=FALSE}
starttime = Sys.time()
debug = FALSE
downloaddata=TRUE
```

@@

thiscolors = c("black", "red")
hist(foldchange_DP_N, breaks=100, freq=F, border=thiscolors[1], main="Reproduced FC for DP vs. N")
hist(foldchange_DP_N[table_s5$Probe], breaks=100, add=T, freq=F, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
```

The reproduced fold changes for the probes listed in Table S5 are clearly more extreme than the rest.

The next test is a test for differentially expressed genes using ANOVA.
>  First, we computed the statistical significance of differential expression between treatments using the Analysis of Variance (ANOVA) method for each probe [21]...<cite> Towfic et al.

```{r}
anova_p = apply(datamatrices[["real_combat_covariates"]], 1,
              function(x){ summary(aov(x[sampleannotation$covariate %in% c("DP", "N")]~
                                         sampleannotation$covariate[sampleannotation$covariate %in% c("DP", "N")]))[[1]][1,5]})

par(mfrow=c(1, 2))
plot(table_s5[,"anova_p"], anova_p[table_s5$Probe], xlab="Original Table S5", 
     ylab="Reproduced", main="ANOVA P-values from table S5 vs. reproduced")
thiscolors = c("black", "red")
hist(anova_p, breaks=100, freq=T, border=thiscolors[1], main="Reproduced ANOVA p-values for DP vs. N")
hist(anova_p[table_s5$Probe], breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
```

@@

```{r dev='png', fig.width=10, fig.height=6, fig.show='hold', tidy=FALSE}
cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
limma_ret_alt = eBayes(fit2)
limma_p_alt = limma_ret_alt$p.value
names(limma_p_alt) = names(limma_ret_alt$Amean)
par(mfrow=c(1, 2))
plot(table_s5[,"limma_p"], limma_p_alt[table_s5$Probe], xlab="Original Table S5", 
     ylab="Reproduced", main="LIMMA P-values from table S5 vs. reproduced")
thiscolors = c("black", "red")
hist(limma_p_alt, breaks=100, freq=T, border=thiscolors[1], main="Reproduced LIMMA p-values for DP vs. N")
hist(limma_p_alt[table_s5$Probe], breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
```
> In parallel, we used Comparative Marker Selection as implemented in GenePattern [26] to directly compare probes between generic and GA. We applied two separate techniques within this framework; a traditional T-test and a Signal-to-Noise Ratio test (SNR).

The T-test is straight forward.
```{r dev='png', fig.width=10, fig.height=6, fig.show='hold', tidy=FALSE}
ttest_p = apply(datamatrices[["real_combat_covariates"]], 1, 
                function(x) { t.test(x[sampleannotation$covariate=="DP"], 
                                     x[sampleannotation$covariate=="N"])$p.value } )
par(mfrow=c(1, 2))
plot(table_s5[,"t_test_p"], ttest_p[table_s5$Probe], xlab="Original Table S5", 
     ylab="Reproduced", main="t-test P-values from table S5 vs. reproduced")
thiscolors = c("black", "red")
hist(ttest_p, breaks=100, freq=T, border=thiscolors[1], main="Reproduced t-test p-values for DP vs. N")
hist(ttest_p[table_s5$Probe], breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
```

> ...we used the Wilcoxon Rank Sum Test [27] as implemented in R ...

```{r dev='png', fig.width=10, fig.height=6, fig.show='hold', tidy=FALSE}
# wilcox.test gives a few hundred warnings for ties:
# Warning: cannot compute exact p-value with ties
wilcoxon_p = suppressWarnings(
  apply(datamatrices[["real_combat_covariates"]], 1, 
        function(x) { wilcox.test(x[sampleannotation$covariate=="DP"],
                                  x[sampleannotation$covariate=="N"])$p.value } ))
par(mfrow=c(1, 2))
plot(table_s5[,"wilcoxon_p"], wilcoxon_p[table_s5$Probe], xlab="Original Table S5", 
     ylab="Reproduced", main="wilcoxon P-values from table S5 vs. reproduced")
thiscolors = c("black", "red")
hist(wilcoxon_p, breaks=100, freq=T, border=thiscolors[1], 
     main="Reproduced wilcoxon p-values for DP vs. N")
hist(wilcoxon_p[table_s5$Probe], breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)

@@

Table S5 on the other hand consists of results from significance tests that are strongly influenced by ComBat's data transformation. An alternative way of handling batch effect compared to ComBat's is to include it in the statistical test. This can be done in ANOVA.

```{r tidy=FALSE}

DP = sampleannotation$covariate=="DP"
N = sampleannotation$covariate=="N"
thisdata =datamatrices[["real_qnorm"]]
cov_fac = factor(sampleannotation$covariate[DP|N])
chip_fac = factor(sampleannotation$chip[DP|N])
#anova_p1 = apply(thisdata, 1, function(x){ summary(aov(x[DP|N]~chip_fac*cov_fac))[[1]][2,5]}  ) 
anova_woc = apply(thisdata, 1, function(x){ summary(aov(x[DP|N]~chip_fac+cov_fac))[[1]][2,5]}  ) 
#co = list(cov_fac=contr.sum, chip_fac=contr.sum)
#anova_p3 = apply(thisdata, 1, function(x){ Anova(lm(x[DP|N] ~  cov1_fac + chip_fac,  contrasts=co), type=3)[[4]][2]}  ) 
table(p.adjust(anova_p , method="fdr")<0.05, p.adjust(anova_woc , method="fdr")<0.05, dnn=c("ComBat adj", "ANOVA adj"))
```
The number of significant probes using the given FDR threshold of 0.05 when batch effect is handled by ANOVA is 
**`r sum(p.adjust(anova_woc , method="fdr")<0.05)`**. This is much lower compared to when ComBat is applied, 
**`r sum(p.adjust(anova_p , method="fdr")<0.05)`**.  The p-value distribution is still skewed, but not as much.
```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
thiscolors = c("black", "red")
hist(anova_p, breaks=100, freq=T, border=thiscolors[1], main="Reproduced ANOVA p-values for DP vs. N")
hist(anova_woc, breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("ComBat adjusted", "ANOVA adjusted"), text.col=thiscolors)
```

LIMMA also has the possibility to handle batch inside the statistical test.
```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
group = factor(sampleannotation$covariate)
block = as.factor(sampleannotation$chip)
design = model.matrix(~0+group+block)
fit = lmFit(thisdata, design)
cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
limma_ret_woc = eBayes(fit2)
limma_p_woc = limma_ret_woc$p.value
names(limma_p_woc) = names(limma_ret_woc$Amean)
rm(design, group, block, fit, cont.matrix, fit2)
table(p.adjust(limma_p_alt , method="fdr")<0.05, p.adjust(limma_p_woc , method="fdr")<0.05, dnn=c("ComBat adj", "LIMMA adj"))
```
The number of significant probes using the given FDR threshold of 0.05 when batch effect is handled by LIMMA is 
**`r sum(p.adjust(limma_p_woc , method="fdr")<0.05)`**. This is much lower compared to when ComBat is applied, 
**`r sum(p.adjust(limma_p_alt , method="fdr")<0.05)`**.  The p-value distribution is still skewed, but not as much.
```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
hist(limma_p_alt, breaks=100, freq=T, border=thiscolors[1], main="Reproduced ANOVA p-values for DP vs. N")
hist(limma_p_woc, breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("ComBat adjusted", "LIMMA adjusted"), text.col=thiscolors)
```

The T-test can not handle batch inside the test. An alternative way of performing a t-test is to subset the data and do the comparison within each batch. For each of the 7 batches that have both a DP ("GA") and a N ("generic") sample, we can do a direct comparison and in effect ending up with 7 samples. Only 7 samples will give less power but the caveats because of batch effect and adjustments are also far less.
```{r tidy=FALSE, results='hold'}

# subsetting the samples so that they have a balanced conditon distribution across batches. i.e each batch has the same compitition of conditions (example 70% treated vs 30% untreated)
drawbatchbalanceddsamples = function(samplenames, batches, conditions)
{
  # 1. find the minimal count for each condition across batches.
	# 2 find the minimal given count / minimal count inside each batch
	# 3 If greater than 1, scale up all conditions for that batch.
	# 4 draw samples randomly from the above calculated condition-count for each batch
	# 5 create and return the data subset.

	#1
	starttab = table( data.frame(conditions, batches ))	
	tab = starttab[, apply(starttab, 2,  min)>0] # do not use batches with missing conditions
	if(dim(tab)[2]==0)
	{
		warning(  paste("Unable to find a balanced subset that has at least one of each condition. Remove one or more conditions and try again")) 
		return(vector())
	}
	if(dim(tab)[1]<dim(starttab)[1])
	{
		warning(  paste("Lost batches during balancing due to missing conditions in those. ",   "Tried: ", paste( dimnames(starttab)[[2]], collapse=" "), "Got:", paste( dimnames(tab)[[2]], collapse=" "  ) ))
	}

	mincond = apply(tab, 1,  min)

	#2
	tab2 = mincond/tab

	#3
	scalings = apply( 1/tab2, 2, min)
	scaledtab = tab
	scaledtab[,] = mincond	
	scaledtab = t(t(scaledtab) * scalings)
	scaledtab = round(scaledtab,0)

	winnersamples= vector()
	for(c in dimnames(scaledtab)[[1]])
	{
		for(b in dimnames(scaledtab)[[2]])
		{
			a = batches==b & conditions==c
			winnersamples = c( winnersamples, sample( samplenames[a], scaledtab[c,b]))
		}
	}
	return(winnersamples)	
}

a = sampleannotation$covariate %in% c("N", "DP")
nbalanceddatasets = 3
pairedpvals = list()
for(i in 1:nbalanceddatasets)
{
  bs = drawbatchbalanceddsamples(sampleannotation$code[a], sampleannotation$chip[a], sampleannotation$covariate[a])
  print(table(sampleannotation[bs,c("chip", "covariate")])) 
  bs = bs[order(sampleannotation[bs, "covariate"], sampleannotation[bs, "chip"])]
  sampleannotation[bs, c("covariate", "chip")] # the samples are ordered
  thispaireddata = datamatrices[["real_qnorm"]][,bs][,1:7] -  datamatrices[["real_qnorm"]][,bs][,8:14]
  pairedpvals[[i]] = apply(thispaireddata, 1, 
                function(x) { t.test( x)$p.value } )
  print(table(p.adjust(pairedpvals[[i]] , method="fdr")<0.05))
  rm(bs,thispaireddata)  
}
```

```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
par(mfrow=c(2, 2))
hist(ttest_p, breaks=100, freq=T, border="black", main="P-values t-test on ComBat adjusted data")
for(i in 1:length(pairedpvals))
{
  hist(pairedpvals[[i]], breaks=100, freq=T, border="black", main=paste("P-values paired t-test draw ", i, sep=""))
}

@@

```{r}
set.seed(100)
datamatrices[["random_raw"]] = matrix(rnorm(length(datamatrices[["real_raw"]])), 
                                      nrow=nrow(datamatrices[["real_raw"]]), 
                                      ncol=ncol(datamatrices[["real_raw"]]), 
                                      dimnames=dimnames(datamatrices[["real_raw"]]))
datamatrices[["random_combat_covariates"]]= as.matrix(ComBat(dat=datamatrices[["random_raw"]],
                                                             batch=sampleannotation$chip, 
                                                             mod=combatmod, 
                                                             numCovs=NULL, 
                                                             par.prior=TRUE, 
                                                             prior.plots=FALSE))
```

```{r}
group = factor(sampleannotation$covariate)
design = model.matrix(~0 + group)
fit = lmFit(datamatrices[["random_combat_covariates"]], design)
cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
limma_ret_rand = eBayes(fit2)
limma_p_rand = limma_ret_rand$p.value
names(limma_p_rand) = names(limma_ret_rand$Amean)
table(p.adjust(limma_p_rand, method="fdr")<0.05)
```

```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
par(mfrow=c(1, 1))
thiscolors = c("black", "red")
hist(limma_p_alt, breaks=100, freq=T, border=thiscolors[1], main="LIMMA p-values for DP vs. N")
hist(limma_p_rand, breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("Reproduced from Real data", "Random data ComBat adjusted"), text.col=thiscolors)
```

```{r}

anova_p_rand = apply(datamatrices[["random_combat_covariates"]], 1,
              function(x){ summary(aov(x[sampleannotation$covariate %in% c("DP", "N")]~
                                         sampleannotation$covariate[sampleannotation$covariate %in% c("DP", "N")]))[[1]][1,5]})
table(p.adjust(anova_p_rand, method="fdr")<0.05)
```

```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
par(mfrow=c(1, 1))
thiscolors = c("black", "red")
hist(anova_p, breaks=100, freq=T, border=thiscolors[1], main="ANOVA p-values for DP vs. N")
hist(anova_p_rand, breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("Reproduced from Real data", "Random data ComBat adjusted"), text.col=thiscolors)
```

@@

Another sanity check is by permuting the DP and N labels and look how the result changes. In order to retain the batch effect we will only swap labels within a batch. This should in theory give fewer significant probes.

```{r}
source("../../commonscripts/helper_functions.r")
#shufflesamplesinbatch = function(samplenames, batch, covariate, shufflecovariates)
#{
  #ret = samplenames
	#b = covariate %in% shufflecovariates
	#for(thisbatch in unique(batch))
	#{
	#	a = batch== thisbatch		
#		ret[a & b] = sample( samplenames[a&b],sum(a&b),  replace=FALSE)		
	#}
	#return(ret)
#}

nshuffleddatasets = 10
datamatrices_permuted = list()
for(i in 1:nshuffleddatasets)
{
  x = shufflesamplesinbatch( sampleannotation$code, sampleannotation$chip, sampleannotation$covariate, c("N", "DP") )
  print(  sum(sampleannotation[,"covariate"] != sampleannotation[x,"covariate"]  ))  # see how many got change. Has 11 N's)  
  datamatrices_permuted[[i]] = as.matrix(ComBat(dat=datamatrices[["real_qnorm"]][, x], 
                                                batch=sampleannotation$chip, 
                                                mod=combatmod, 
                                                numCovs=NULL, 
                                                par.prior=TRUE, 
                                                prior.plots=FALSE))
}

permuted_pvalcounts=list()
for(i in 1:nshuffleddatasets)
{ 
  group = factor(sampleannotation$covariate)
  design = model.matrix(~0 + group)
  fit = lmFit(datamatrices_permuted[[i]], design)
  cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
  fit2 = contrasts.fit(fit, cont.matrix)  
  this_p = eBayes(fit2)$p.value
  permuted_pvalcounts[[i]] = hist(this_p, plot=FALSE, breaks=100)$counts
  print(table(p.adjust(this_p, method="fdr")<0.05))
}
```

```{r, dev='svg', fig.width=8, fig.height=8, tidy=FALSE}
real_pvalcounts=hist(limma_p_alt, plot=FALSE, breaks=100)$counts
plot((1:20)/100, real_pvalcounts[1:20], col="red",
     main="P-values, permuted real data", 
      xlab="p-value", ylab="Frequency", type="l", lwd=2, 
      ylim=c(0, max(unlist( c(permuted_pvalcounts,real_pvalcounts)))))
for(i in 1:length(permuted_pvalcounts))
{
  lines((1:20)/100,permuted_pvalcounts[[i]][1:20], col="blue")
}
legend("topright", legend=c("Reproduced p-values real labels", "Permuted DP and N labels"), text.col=c("red", "blue"))
```

Forslag figur brukt i artikkel.
```{r pvaluestowfic, dev='svg', fig.width=10, fig.height=10, tidy=FALSE}
# figure used in our article
# NB! bruker orgniale data som ikke er floored eller log2!
par(mfrow=c(2, 2))

@@

# random ComBat + limma
hist(limma_p_rand, main="P-values, Random data, ComBat followed by Limma", breaks=100, xlab="p-value" , ylim=thisylim)

@@

>>>>

This document has five main parts
- Getting and formatting the data
- Reproduce some of the results to show that we are working on the same data and analysis workflow
- Remove the use of ComBat and perform a similar analysis with alternative established tools
- Estimate the error introduced by ComBat and the consequences for the conclusion of the study
- Perform a few more sanity checks to substantiate that the difference in results for the two above analysis is mainly introduced by ComBat

```{r runparameters, results='hide', echo=FALSE}
starttime = Sys.time()
debug = TRUE
downloaddata=TRUE
set.seed(100)
```

```{r, results='hide', message=FALSE}
includelibs = c("Biobase", "GEOquery", "sva", "limma",
                 "preprocessCore")
lapply(includelibs, require, character.only=T)
```

@@

thiscolors = c("black", "red")
hist(foldchange_DP_N, breaks=100, freq=F, border=thiscolors[1], main="Reproduced FC for DP vs. N")
hist(foldchange_DP_N[table_s5$Probe], breaks=100, add=T, freq=F, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
```
The first plot reveals that the reproduced fold changes are much more compressed than the ones in Table S5.
There seems to two linear relationship, one for negative FC's and another for positive FC's. They seems to have different intercepts but comparable slope. The second plot shows that the reproduced fold changes for the probes listed in Table S5 are clearly more extreme than the rest.

The fold changes in table 5S do not match the average expression values listed in table S2 either as a the same plot will show;

```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
# table S2 has the AVG for both DP and N listed
fc_table_s2 = table_s2$AVG_GA - table_s2$AVG_generic
names(fc_table_s2) = table_s2$ID
x = table_s5$Fold_Change
y = fc_table_s2[table_s5$Probe]
lim =c( min(   c(min(x, na.rm=T), min(y, na.rm=T))  ), max( c(max(x, na.rm=T), max(y, na.rm=T)) ))
plot(x, y , ylim=lim, xlim=lim, pch="." ,  xlab="Table_S5", ylab="Table_S2",
     main="FC comparison, given values.") #  almost perfect match with offset.
rm(x, y, lim)
```
There seems to be some inconsistancy between between the fold changes in table S5 and the reproduced ones, however it is the conficende i.e p-values that are affected by the use of ComBat.

The rest of table S5 is p-values and we will try to reproduce the ones from Limma.

@@

```{r dev='png', fig.width=10, fig.height=6, fig.show='hold', tidy=FALSE}
cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
limma_ret_alt = eBayes(fit2)
limma_p_alt = limma_ret_alt$p.value
names(limma_p_alt) = names(limma_ret_alt$Amean)
par(mfrow=c(1, 2))
plot(table_s5[,"limma_p"], limma_p_alt[table_s5$Probe], xlab="Original Table S5", 
     ylab="Reproduced", main="LIMMA P-values from table S5 vs. reproduced")
thiscolors = c("black", "red")
hist(limma_p_alt, breaks=100, freq=T, border=thiscolors[1], main="Reproduced LIMMA p-values for DP vs. N")
hist(limma_p_alt[table_s5$Probe], breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("All probes", "Table S5 probes"), text.col=thiscolors)
```

In addition to the Limma test, Towfic et al. performed similar tests for significant probes between generic and GA using several other methods. 
We tried several of them as well, but have skipped them in this report since they showed a similar patterent as for the Limma p-values.

@@

Table S5 on the other hand consists of results from significance tests that are strongly influenced by ComBat's data transformation. An alternative way of handling batch effect compared to ComBat's is to include it in the statistical test. This can be done in Limma

```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
group = factor(sampleannotation$covariate)
block = as.factor(sampleannotation$chip)
design = model.matrix(~0+group+block)
fit = lmFit(thisdata, design)
cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
fit2 = contrasts.fit(fit, cont.matrix)
limma_ret_woc = eBayes(fit2)
limma_p_woc = limma_ret_woc$p.value
names(limma_p_woc) = names(limma_ret_woc$Amean)
rm(design, group, block, fit, cont.matrix, fit2)
table(p.adjust(limma_p_alt , method="fdr")<0.05, p.adjust(limma_p_woc , method="fdr")<0.05, dnn=c("ComBat adj", "LIMMA adj"))
```
The number of significant probes using the given FDR threshold of 0.05 when batch effect is handled by LIMMA is 
**`r sum(p.adjust(limma_p_woc , method="fdr")<0.05)`**. This is much lower compared to when ComBat is applied, 
**`r sum(p.adjust(limma_p_alt , method="fdr")<0.05)`**.  The p-value distribution is still skewed, but not as much.
```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
hist(limma_p_alt, breaks=100, freq=T, border=thiscolors[1], main="Reproduced ANOVA p-values for DP vs. N")
hist(limma_p_woc, breaks=100, add=T, freq=T, border=thiscolors[2])
legend("topright", legend=c("ComBat adjusted", "LIMMA adjusted"), text.col=thiscolors)
```

To substantiate that the result from the use of ComBat is less trustworthy than the alternative analysis we provide a couple of additional sanity checks.

First we use random numbers drawn from the same distribution(mean=0, sd=1) regardless of batch or covariate but retaining the batch/covariate design.
```{r}
set.seed(100)
datamatrices[["random_raw"]] = matrix(rnorm(length(datamatrices[["real_raw"]]), mean=0, sd=1), 
                                      nrow=nrow(datamatrices[["real_raw"]]), 
                                      ncol=ncol(datamatrices[["real_raw"]]), 
                                      dimnames=dimnames(datamatrices[["real_raw"]]))
datamatrices[["random_combat_covariates"]]= as.matrix(ComBat(dat=datamatrices[["random_raw"]],
                                                             batch=sampleannotation$chip, 
                                                             mod=combatmod, 
                                                             numCovs=NULL, 
                                                             par.prior=TRUE, 
                                                             prior.plots=FALSE))
```

Limma is then used in 3 ways
- On the ComBat adjusted random numbers
- On the random numbers ignoring batch information
- On the random numbers including batch as a blocking factor

```{r}
group = factor(sampleannotation$covariate)
design = model.matrix(~0 + group)
cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)

fit = lmFit(datamatrices[["random_combat_covariates"]], design)  
fit2 = contrasts.fit(fit, cont.matrix)
limma_p_rand_combat = eBayes(fit2)$p.value[,1]

fit = lmFit(datamatrices[["random_raw"]], design)  
fit2 = contrasts.fit(fit, cont.matrix)
limma_p_rand_nocombat = eBayes(fit2)$p.value[,1]

block = as.factor(sampleannotation$chip)
design = model.matrix(~0+group+block)
cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)
fit = lmFit(datamatrices[["random_raw"]], design)  
fit2 = contrasts.fit(fit, cont.matrix)
limma_p_rand_batchblocked = eBayes(fit2)$p.value[,1]

@@

```{r dev='png', fig.width=8, fig.height=8, fig.show='hold', tidy=FALSE}
par(mfrow=c(1, 1))
thiscolors = c("red", "blue", "black")
hist(limma_p_rand_combat, breaks=100, freq=T, border=thiscolors[1], main="P-values, Random numbers for DP vs. N")
hist(limma_p_rand_nocombat, breaks=100, add=T, freq=T, border=thiscolors[2])
hist(limma_p_rand_batchblocked, breaks=100, add=T, freq=T, border=thiscolors[3])
legend("topright", legend=c("ComBat adjusted", "Limma adjusted", "No adjustment"), text.col=thiscolors)
```

@@

Another sanity check is by permuting the DP (GA) and N(generic) labels and observe changes in the results. In order to retain the batch effect we will only swap labels within a batch. This should in theory give fewer significant probes.

```{r results='hide'}
source("../../commonscripts/helper_functions.r")

nshuffleddatasets = 10
datamatrices_permuted = list()
changedlabels=vector()
for(i in 1:nshuffleddatasets)
{
  x = shufflesamplesinbatch( sampleannotation$code, sampleannotation$chip, sampleannotation$covariate, c("N", "DP") )
  changedlabels[i] = (  sum(sampleannotation[,"covariate"] != sampleannotation[x,"covariate"]  ))  # see how many got change. Has 11 N's)  
  datamatrices_permuted[[i]] = as.matrix(ComBat(dat=datamatrices[["real_qnorm"]][, x], 
                                                batch=sampleannotation$chip, 
                                                mod=combatmod, 
                                                numCovs=NULL, 
                                                par.prior=TRUE, 
                                                prior.plots=FALSE))
}
```
**`r nshuffleddatasets`** permutations were created and batch adjusted with ComBat.

Then perform the limma test on the ComBat adjusted data as was done for the real data.
```{r}
permuted_pvalcounts=list()
overviewtab = data.frame(permutation=1:nshuffleddatasets, changedlabels=changedlabels, significantprobes=NA)
for(i in 1:nshuffleddatasets)
{ 
  group = factor(sampleannotation$covariate)
  design = model.matrix(~0 + group)
  fit = lmFit(datamatrices_permuted[[i]], design)
  cont.matrix = makeContrasts ( contrasts="groupDP-groupN", levels=design)  
  fit2 = contrasts.fit(fit, cont.matrix)  
  this_p = eBayes(fit2)$p.value
  permuted_pvalcounts[[i]] = hist(this_p, plot=FALSE, breaks=100)$counts 
  overviewtab$significantprobes[i] =sum(p.adjust(this_p, method="fdr")<0.05)
}
print(overviewtab)
rm(overviewtab, group, design, fit, cont.matrix, fit2, this_p)
```
The table lists the number of labels that are changed compared to the real ones for each permutation. Remember that the totat number of N (generic) samples is **`r sum(sampleannotation$covariate=="N")`**. So even when about half of those are swapped with DP(GA), the ComBat adjusted data produces many significant genes. The number of significant genes using the same 0.05 FDR cut-off for the real labels was **`r sum(p.adjust(limma_p_alt , method="fdr")<0.05)`**.

The P-value distribution can slo be plotted.
```{r, dev='svg', fig.width=8, fig.height=8, tidy=FALSE}
real_pvalcounts=hist(limma_p_alt, plot=FALSE, breaks=100)$counts
plot((1:20)/100, real_pvalcounts[1:20], col="red",
     main="P-values, permuted real data", 
      xlab="p-value", ylab="Frequency", type="l", lwd=2, 
      ylim=c(0, max(unlist( c(permuted_pvalcounts,real_pvalcounts)))))
for(i in 1:length(permuted_pvalcounts))
{
  lines((1:20)/100,permuted_pvalcounts[[i]][1:20], col="blue")
}
legend("topright", legend=c("Reproduced p-values real labels", "Permuted DP and N labels"), text.col=c("red", "blue"))
```
Shuffling the N and DP labels whitinh batches produces about the same p-value distribution as the real data. This is an indication that the significant genes found in the real data is not associated with the N and DP labels, but a rather an effect of the ComBat adjustments.

Forslag 1 figur i artikkel.
```{r pvaluestowfic, dev='svg', fig.width=10, fig.height=10, tidy=FALSE}
# figure used in our article
# NB! bruker orgniale data som ikke er floored eller log2!
par(mfrow=c(2, 2))

@@

# random ComBat + limma
hist(limma_p_rand_combat, main="P-values, Random data, ComBat followed by Limma", breaks=100, xlab="p-value" , ylim=thisylim)

@@

Forslag 2 figur i artikkel.
```{r pvaluestowfic2, dev='svg', fig.width=10, fig.height=10, tidy=FALSE}
# figure used in our article
# NB! bruker orgniale data som ikke er floored eller log2!
par(mfrow=c(1, 1))
thiscolors = c("red", "blue", "black")
thisylim = c(0, max(unlist( c(permuted_pvalcounts,real_pvalcounts))))
xrange=1:20
xrange=1:25
# reproduced ComBat + limma
 plot((xrange)/100, hist(limma_p_alt, breaks=100, plot=FALSE)$counts[xrange], 
       main="P-values",
       xlab="p-value", 
       ylab="frequency",
       ylim=thisylim,
       type="l",
       lwd=2,
       col=thiscolors[1])

# batch handled in limma
lines((xrange)/100, hist(limma_p_woc, breaks=100, plot=FALSE)$counts[xrange], 
     col=thiscolors[2], lwd=2)

# random ComBat + limma
lines((xrange)/100, hist(limma_p_rand_combat, breaks=100, plot=F)$counts[xrange],
       col=thiscolors[3], lwd=2)

legend("topright", legend=c("Real data, ComBat adjusted",
                            "Real data, batch handled by Limma",
                            "Random data, ComBat adjusted"),
       text.col=thiscolors)
